A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-13 linear.tex
rough cut. still missing last discussion section and finite ts VSI results

\section{Linear problem for perfectly coupled dust}\label{linear_problem}
We calculate explicit solutions to the linearized
equations in the limit of perfectly coupled dust ($\tstop=0$), so that
the equilibria defined in \S\ref{eqm} are exact steady states. We 
consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta f (r, z) = \delta f_1(r,z)\exp{(\ii k r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k$ is a real wavenumber such that $|kr|\gg 1$, and the
amplitude $\dd f_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

Introducing 
\begin{align}
  W \equiv \frac{\dd\rho}{\rho}, \quad Q \equiv \frac{\dd P}{\rho},
\end{align}
the linearized equations for 
locally isothermal, perfectly-coupled dusty gas with the pressure
equation in place of the dust-fraction
(Eq. \ref{masseq}---\ref{momeq}, Eq. \ref{eff_energy}) are then:

\begin{align}
  \ii\sigma W &= \ii k \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi + 
  \delta\bm{F}\cdot\hat{\bm{r}},\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &=  \delta\bm{F}\cdot\hat{\bm{z}},\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2}, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
%               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy} 
\end{align}  
where $^\prime \equiv \p_z$ and $\dd\bm{F}$ is the linearized pressure
force, given in Appendix \ref{lin_press}. Note that we have assumed a
temperature profile that only depends on $r$.

@@

Eq. \ref{lin_mass}---\ref{lin_energy} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained.  Two boundary conditions are needed for the $\tstop=0$
problem. This is expected since this problem is exactly equivalent to
 adiabatic hydrodynamics \citep[e.g.][]{lubow93}. For simplicity
we impose solid boundaries so that $\delta v_z(\pm\zmax)=0$.

We solve the linearized equations as a generalized eigenvalue problem 
using a pseudo-spectral code adapted from \cite{lin15}. Amplitudes 
are expanded in Chebyshev polynomials up to order $N_z=512$. We check
results using Eq. \ref{vsi_check}.

>>>>

\section{Quasi-global linear problem}\label{linear_problem}
We consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta f (r, z) = \delta f_1(r,z)\exp{(\ii k r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k$ is a real wavenumber such that $|kr|\gg 1$, and the
amplitude $\dd f_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

Introducing 
\begin{align}
  W \equiv \frac{\dd\rho}{\rho}, \quad Q \equiv \frac{\dd P}{\rho},
\end{align}
the linearized equations for 
locally isothermal dusty gas with the pressure
equation in place of the dust-fraction
(Eq. \ref{masseq}---\ref{momeq}, Eq. \ref{eff_energy}) are then:

\begin{align}
  \ii\sigma W &= \ii k \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k Q,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right]  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy} 
\end{align}  
where $^\prime \equiv \p_z$ and $\bm{F} \equiv -\nabla P/\rho$. 
The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
Note that we have assumed a temperature profile that only depends on $r$.

@@

Eq. \ref{lin_mass}---\ref{lin_energy} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. Numerical solutions are generally required for the 
verticall-stratified problem, as we pursue below.

However, in the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain a local dispersion relation. This is done in 
Appendix \ref{compressible_streaming}. There we show the above linearized
equations give the streaming instability in the appropriate limit 
\citep{youdin05a,jacquet11}.

### 2016-08-13 model.tex
more discussions
rough intro
doing last section
rough cut. still missing last discussion section and finite ts VSI results

\subsection{Locally isothermal equation of state} 
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \tepsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time. 
In numerical calculations we will consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

@@

\section{Limiting behaviours}

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations about an axisymmetric steady state. A variable $f$ is
subject to perturbations of the form 
\begin{align}
  \delta f(r,z)\exp{\left(-\ii\sigma t\right)},
\end{align}
where $\sigma$ is the complex mode frequency. In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real coefficients $A,B,C,D$ can be read off 
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}). We now consider various limits of
Eq. \ref{int_rel}.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
%\begin{align}
%  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
%  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
%    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
%\end{align} 
\begin{align}
  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the 
square of the epicylic frequency, and we assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} has the identical form
to the Solberg-Hoiland criteria for adiabatic gas dynamics 
\citep[e.g.][]{tassoul78}, except here the entropy $s$ is really a
representation of the dust distribution via $s=\ln{\left[c_s^2\left(1
    - \tepsilon\right)\right]}$.

@@

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
When $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel}
indicates that $\sigma^2$ is generally complex, implying instability.  
Writing $\sigma = 
-\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
growth rates, respectively, we have 
\begin{align}
  s = \frac{\imag\int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check}
\end{align} 
This is classic VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We compute
explicit solutions in this limit in
\S\ref{linear_problem}---\ref{results}.

@@

\subsection{Strictly isothermal gas well-coupled to dust} 
Now consider constant $c_s$ but $\tstop\neq0$. Then Eq. \ref{int_rel}
gives 
\begin{align}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, 
\end{align}
%In this case 
%the effective energy equation is 
%\begin{align}\label{dusty_cooling}
%  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
%\end{align} 
%The function $\mathcal{C}$ can be considered as a cooling term. 
%By
%analogy with enabling VSI in an adiabatic gas through cooling
%\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
%if this term is large enough (physically meaning diffusion of the
%dust-fraction is strong). 
%Note that $\mathcal{C}$ is also responsible for the streaming
%instability  in the strong-drag limit \citep{laibe14}, though this
%does not require vertical stratification. In the form of
%Eq. \ref{dusty_cooling}, however, the streaming instability has a
%thermodynamical interpretation, similar to thermal instabilities in
%stars \citep{latter06}.  
which represent instabilities due to large-but-finite dust-gas
friction. Our formalism gives these dusty instabilities a 
thermodynamical interpretation because the dust-gas friction enters the problem
as an energy source term \citep[cf. thermal instabilities in
  stars, see also][]{latter06}. 
This includes the streaming instability in
the strong-drag limit \citep{youdin05a, jacquet11, laibe14}.

>>>>

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \tepsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time. 
In numerical calculations we will consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

@@

\section{Limiting behaviours}\label{limits}

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations about an axisymmetric steady state. A variable $f$ is
subject to perturbations of the form 
\begin{align}
  \delta f(r,z)\exp{\left(-\ii\sigma t\right)},
\end{align}
where $\sigma$ is the complex mode frequency. In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real coefficients $A,B,C,D$ can be read off 
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}). We now consider various limits of
Eq. \ref{int_rel}.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the 
square of the epicylic frequency. We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy in the standard
expression for said criteria \citep[e.g.][ where $C_p$ is the heat
  capacity]{tassoul78}.

%except here the entropy $s$ is really a
%representation of the dust distribution via $s=\ln{\left[c_s^2\left(1
%    - \tepsilon\right)\right]}$.

@@

@@

\subsection{Thermodynamic view of dusty instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
%When $c_s(r,z)$ is non-uniform but $\tstop=0$, 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations. Writing $\sigma = -\omega + \ii s$
where $\omega$ and $s$ are the real frequency and growth rates, 
respectively, the imaginary part of Eq. \ref{int_rel} gives 
\begin{align}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
\end{align}
%In this case 
%the effective energy equation is 
%\begin{align}\label{dusty_cooling}
%  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
%\end{align} 
%The function $\mathcal{C}$ can be considered as a cooling term. 
%By
%analogy with enabling VSI in an adiabatic gas through cooling
%\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
%if this term is large enough (physically meaning diffusion of the
%dust-fraction is strong). 
%Note that $\mathcal{C}$ is also responsible for the streaming
%instability  in the strong-drag limit \citep{laibe14}, though this
%does not require vertical stratification. In the form of
%Eq. \ref{dusty_cooling}, however, the streaming instability has a
%thermodynamical interpretation, similar to thermal instabilities in
%stars \citep{latter06}.  
which represent instabilities due to dust-gas friction. 
%Our formalism
%gives these dusty instabilities a thermodynamical interpretation. 
% We further discuss this concept in \S\ref{dust_as_thermo}. 
%In \S\ref{dust_work} we showed that instabilities may 
%arise from the term proportional $\imag\left( \nabla\cdot\dd\bm{v}^*
%\delta C\right)$. 
 Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

It is well-known that such correlations can lead to pulsational
instabilities in stars \citep{cox67}. We thus suggest an interpretation
of dust-drag instabilities in disks in a similar way. We give a qualitative
description below, based on the more rigorous treatments of stellar 
pulsational instabilities by 
\cite{cox67} and lecture notes from \cite{samadi15} and
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/}}.

%and disks
%\cite{kato78}.

Consider the Lagragian perturbation $\Delta$ of a variable $f$, 
related to the Eulerian perturbation by $\Delta f = \delta f + 
\bm{\xi}\cdot\nabla f$, where $\bm{\xi}$ is the Lagragian  
displacement ($\dd v_{x,z} = -\ii\sigma\xi_{x,z}$). %Note that $\Delta
%\mathcal{C} = \delta \mathcal{C}$ for a uniform background $C$ as
%required for equilibrium. 
Then Eq. \ref{lin_mass_full}---\ref{lin_energy_full} imply  
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). 
\end{align}
This reflects `$TdS  =  PdV$' from thermodynamics, if the 
internal energy is constant. That is, dust-gas drag provides 
heating ($TdS$) to do $PdV$ work. Whether positive or negative work is 
done depends on the correlation between the pressure and density
perturbations, as follows.

The full energy equation is 
\begin{align*}
  \frac{DP}{Dt} = \frac{P}{\rho}\frac{D\rho}{D t} + \mathcal{C}. 
\end{align*}
Dust-drag causes a phase difference between the 
pressure and density evolution following a parcel. Now, consider a 
fluid element at maximum density during an oscillation cycle, 
so that $D\rho/ D t = 0$. If at this point dust-gas drag provides an
effective heating ($\mathcal{C}>0$), then the fluid element is
experiencing increasing pressure, $DP/D t>0$. It will attain
pressure maximum \emph{after} reaching density maximum. This implies 
pressure maximizes while the parcel is expanding: it will do 
positive work (the enclosed area by its `$P$-$V$' curve  is 
positive). This will increase the oscillation amplitude.

Indeed, \cite{jacquet11} explains the essence of the streaming 
instability in dusty protoplanetary disks as dust accumulation
(and hence compression) at a pressure bump which back-reacts on the
gas flow in such a way to strengthen the pressure bump. A phase
difference was also noted in the numerical calculations of
\cite[][although they consider the full two-fluid system]{youdin07b}.

We suggest that a useful way to understand  
dust-drag instabilities in protoplanetary disks may be to re-cast the  
problem into that of thermodynamics, since such instabilities have 
already been well-studied in the stellar context.

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
If $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel} 
gives  
\begin{align}
  s = \frac{\imag\int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check}
\end{align} 
This is classic VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We compute
explicit solutions in this limit in \S\ref{results}.
